Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-22 paper.tex
Write up compression benchmarks

Fig~\ref{fig-compression-chunksize} shows that chunk size has a 
fairly weak influence on compression ratio. While larger 
sample chunk sizes lead to a slight increase in compression
ratio on the more compressable fields, the overall effect
on total storage used will be slight. Variant chunk size 
appears to have no effect on compression ratio. Thus, we 
do not expect the choice of chunk size to have a significant
impact on overall storage usage, and so choice should be
determined by other considerations such as expected data
access patterns.

[TODO, reffing Fig~\ref{fig-compression-shuffle} and 
Fig~\ref{fig-compression-chunksize}.]

@@

\begin{figure}
\includegraphics{figures/compression-chunksize}
\caption{Effects of chunk sizes on compression ratio on call-level 
fields in 1000 Genomes data.
(A) Varying sample chunk size, holding variant chunk size fixed at 10,000.
(B) Varying variant chunk size, holding sample chunk size fixed at 1,000.
In all cases the zstd compressor with clevel=7 was used. Bit shuffle
was used for call\_genotype, and no shuffle used for the other fields.
\label{fig-compression-chunksize}}
\end{figure}

>>>>

Fig~\ref{fig-compression-chunksize} shows that chunk size has a 
weak influence on compression ratio for most fields. Interestingly,
the choice of chunk size has a somewhat unpredictable effect
on compression for the genotype matrix, suggesting that how 
samples are grouped together has a strong effect on compression.
At around 50X, compression is excellent in all cases however, 
and given that genotypes contribute relatively little to the
overall storage cost in many datasets 
(Table~\ref{tab-genomics-england-data}), the differences are slight.
Otherwise, increasing sample chunk size slightly increases 
compression on call\_AB, and has no effect on less compressible
fields. Variant chunk size appears to have almost no effect on compression 
ratio. Thus, we do not expect the choice of chunk size to have a significant
impact on overall storage usage, and so choice should be
determined by other considerations such as expected data
access patterns.

The ``shuffle'' parameter in the Blosc meta-compressor~\cite{alted2010modern}
can result in substantially better compression, albeit at the 
cost of somewhat slower decoding (see Fig~\ref{fig-whole-matrix-decode}).
Fig~\ref{fig-compression-shuffle} shows the effect of 
the bit shuffle (grouping together bits at the same position
across bytes before compression), and byte shuffle (grouping together 
bytes at the sample position across words before compression)
operations on compression ratio. Bit shuffle provides a significant
improvement in compression for the call\_genotype field  because 
the vast majority of genotype calls will be 0 or 1, and therefore 
bits 1 to 7 will be 0. Thus, grouping these bits together will 
lead to significantly better compression. This strategy also 
works extremely well when compressing boolean fields stored as 
8 bit integers, where the top 7 bits are always 0. In practise,
boolean fields stored in this way have very similar compression
to using a bit-packing pre-compression filter (data not shown).
Although byte shuffle leads to somewhat better compression for call\_AD
and call\_DP, it gives substantially worse compression on call\_AB
than no shuffling. 
The default in \texttt{vcf2zarr}
is therefore to use bit shuffle for \texttt{call\_genotype}
and all boolean fields, and to not use 
byte shuffling on any field. These defaults can be easily
overruled, however, by outputting and modifying a JSON formatted 
storage schema before encoding to Zarr.

@@

\begin{figure}
\includegraphics{figures/compression-chunksize}
\caption{Effects of chunk sizes on compression ratio on call-level 
fields in 1000 Genomes data.
(A) Varying sample chunk size, holding variant chunk size fixed at 10,000.
(B) Varying variant chunk size, holding sample chunk size fixed at 1,000.
In all cases the zstd compressor with clevel=7 was used. Bit shuffle
was used for call\_genotype, and no shuffle used for the other fields.
Values are chosen to be evenly spaced on a linear scale 
between 100 and 2504 (the number of samples) in (A) and 
evenly spaced between 100 and 96514 on a log scale in (B). 
\label{fig-compression-chunksize}}
\end{figure}
